---
title: "Exploring White Wine Quality"
author: "Tobenna Anens"
date: "`r format(Sys.time(), '%d %B, %Y')`"
---

========================================================
```{r echo=FALSE, message=FALSE, warning=FALSE, error=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

# knitr: Suppress code/messages/warnings 
#        Set default plot options and center them
chooseCRANmirror(graphics=FALSE, ind=1)
knitr::opts_chunk$set(fig.width=9,fig.height=5,fig.path='Figs/',
                      fig.align='center',tidy=TRUE, echo=FALSE, warning=FALSE,message=FALSE)

# load packages
library(ggplot2)
library(knitr)
library(GGally)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Load_the_Data}
# Load the Data
setwd('C:/Users/toben/Downloads')
whitewines <- read.csv("wineQualityWhites.csv", header = T)
```

The wineQualityWhites dataset was created in 2009 by Paulo Cortez and other 
contributors. It contains physico-chemical properties and quality ratings for 
nearly 5,000 variants of a white wine.

# Univariate Plots Section

```{r echo=FALSE, warning=FALSE, message=FALSE, Univariate_Plots}
# output the dimensions of whitewines
dim(whitewines)
```

```{r echo=FALSE, warning=FALSE, message=FALSE}
# Output the structure of whitewines
str(whitewines)
```

```{r echo=FALSE, message=FALSE}
# View the first 6 rows of whitewines
head(whitewines)
```

The dataset has 4898 observations and 13 variables. 

```{r echo=FALSE, message=FALSE}
# plot a histogram of quality values
qplot(as.factor(whitewines$quality))
```

A large majority of the wines have middle and upper middle scores (5, 6, 7). 
There are very few with scores of 9 or 3. 

```{r}
# Output the range of values for quality 
range(whitewines$quality)
```

The quality of wines ranges from 3 to 9. That means there are no wines 
with perfect or extremely poor scores in this dataset.


```{r}
# Output a numerical summary for quality
summary(whitewines$quality)
```


```{r}
# See the number of observations with each value of quality
table(whitewines$quality)
```


I wonder what made these 3 and 9 wines stand out and earn their scores? Are the 
wines with 5's 6's and 7's consistent in their properties? Next I'll look at  distributions of all inputs, starting with fixed.acidity 
and volatile.acidity.

```{r, echo=FALSE, message=FALSE}
# plot a histogram of fixed.acidity values
qplot(whitewines$fixed.acidity)
```

Most wines have a fixed acidity between 6 and 7.5.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# plot a histogram of volatile.acidity.
# Limit the x axis to focus on the main body of the data
qplot(whitewines$volatile.acidity, xlim=c(0,0.6))
```

Most wines fall between a volatile acidity of 0.2 and 0.3. 

Is acidity consistent among wines with higher quality? Does it vary by quality 
in a consistent way(Do higher quality wines usually have higher acidity)? I 
expect the "acidity" variables to be correlated with pH.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# plot a histogram of citric.acid. Limit the x-axis for a closer study
qplot(whitewines$citric.acid, xlim=c(0,0.75))
```

Most wines have a citric acid level between 0.2 and 0.4.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Same histogram of citric.acid, but change the number of bins for more 
# granularity
qplot(whitewines$citric.acid, xlim=c(0,0.75), bins=60)
```

Same plot with more bins. A notable amount having a citric acid level of nearly 
0.5. 

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# plot a histogram of residual.sugar. Limit the x-axis, increase number of bins
# for closer study
qplot(whitewines$residual.sugar, xlim=c(0,20), bins=60)
```

The most common residual sugar level for wines is between 1 and 2.5, while there
are a significant number of the wines at higher levels of residual sugar, though 
the number gradually decreases as the sugar level increases.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Perform a log transform of the x-axis for closer study
qplot(whitewines$residual.sugar, bins=60) +
  scale_x_log10()
```

A log transform of the x-axis reveals a bimodal distribution of residual sugar, 
with peaks a little below 1.5 and around 10. I wonder what the average quality 
is for these two peaks? What else makes these two peaks different? 

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# plot a histogram of chlorides. Limit the x-axis for closer study.
qplot(whitewines$chlorides, xlim=c(0,0.1))
```

Most wines have chloride levels between 0.025 and 0.0625.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# plot a histogram of free.sulfur.dioxide. Limit the x-axis for closer study
qplot(whitewines$free.sulfur.dioxide, xlim = c(0,100))
```

Most wines have a free sulfur dioxide content between about 10 and about 60.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# plot a histogram of free.sulfur.dioxide. perform a log transform of the x-axis
# for closer study
qplot(whitewines$free.sulfur.dioxide) +
  scale_x_log10()
```

A log transform changes the distributions' shape a little; the data now appears
more left skewed.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# plot a histogram of total.sulfur.dioxide. Increase the number of bins for
# closer study
qplot(whitewines$total.sulfur.dioxide, bins = 60)
```

Most wines have a total sulfur dioxide content between about 75 and about 225.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# plot a histogram of sulphates. Perform a log transform of the x-axis for 
# closer study
qplot(whitewines$sulphates) +
  scale_x_log10()
```

 
Most wines have sulphates between about 0.3 and and about 0.63. Does sulfur dioxide/sulphate content have a high correlation with quality? Why does the 
sulfur dioxide/sulphate distributions all look right skewed before a log 
transform?

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# plot a histogram of density
qplot(whitewines$density)
```

I will limit the x-axis to study this data more closerly

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# plot a histogram of density. Limit the x-axis for closer study
qplot(whitewines$density, xlim=c(0.98,1.005))
```

Most wines have a density between 0.989 and 0.999. 

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# plot a histogram of alcohol
qplot(whitewines$alcohol)
```

Most wines have alcohol levels between 8.5 and 13.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# plot a histogram of pH
qplot(whitewines$pH)
```

Most wines have a pH between 2.925 and 3.375. 

# Univariate Analysis

### What is the structure of your dataset?
The dataset has 4898 observations of 13 variables. Two variables are qualitative 
(X, which is like an ID; and quality), and the rest are quantitative. The
quality of wines ranges from 3 to 9. Most wines have a quality of 6. The median 
quality is 6 and mean quality is 5.878. 75% of wines have a score of 6 or lower.

### What is/are the main feature(s) of interest in your dataset?
I am most interested in quality. I want to know what makes a wine higher 
quality.

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?
I think the most relevant features for my analysis will be residual sugar, 
alcohol, and acidity/pH. I suspect that higher quality wines will have a
consistent range of values for these features.

### Did you create any new variables from existing variables in the dataset?
I decided to make a rank variable to break the wines into three groups: (3,4] or 
"bad", (4,7] or "average"", and (7,9] or "top". This rank variable helps me to
group wines of similar quality, and study differences between groups of 
differing quality.

```{r, echo=FALSE, message=FALSE}
# Create a variable, rank, by cutting the quality variable in four locations to
# create three buckets
whitewines$rank <- cut(whitewines$quality, breaks = c(2,4,7,9))
```


### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?
The most interesting distribution I noticed was in the residual.sugar plot. 
Prior to any transform, There is a noticable number of wines at every 
level of residual sugar after 5, though it steadily decreases as the level 
increases. But once I applied a log transform on the x axis, I found a bimodal 
distribution!

# Bivariate Plots Section

For my bivariate exploration, I will look at relationships between quality and 
other variables, like fixed and volatile acidity, residual sugar, and alcohol. I
will also look at relationships between supporting variables, like pH and
acidity. I will also look at relationships between pairs that include a variable
that has a strong correlation with quality.

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots, fig.width=12, fig.height=10}
# Create a plot matrix of the first 10 variables
ggpairs(whitewines, axisLabels = "internal", columns = 2:10) +
  theme(text = element_text(size=10), 
     axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1, size=7),
             panel.grid.minor = element_blank(), 
             panel.grid.major = element_blank())  # rotate and resize axis labels
                                                  # and remove grid lines
```

This first plot matrix shows relationships between 9 supporting variables 
(X is excluded. X does not mean much though since X is just an ID). There is a 
positive correlation between residual sugar and density, a negative correlation
between fixed acidity and pH (as expected), an interesting triangle-shaped
distribution for volatile acidity and pH, a strong positive correlation between 
fixed and total sulfur dioxide, a slightly negative correlation between volatile
acidity and citric acid (not expected) and notably strong positive correlation
between fixed acidity and citric acid. 

I will select some of these interesting supporting variables (fixed.acidity:residual.sugar, total.sulfur.dioxide, density and pH) to include 
in a second plot matrix that includes the remaining supporting variables and 
quality.

```{r, echo=FALSE, message=FALSE, fig.width=12, fig.height=10}
# Create a plot matrix of 10 selected variables, including choices of variables
# from the first plot matrix
ggpairs(whitewines, axisLabels = "internal", columns = c(2:5,7,9:13)) +
    theme(text = element_text(size=10), 
     axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1, size=7),
             panel.grid.minor = element_blank(), 
             panel.grid.major = element_blank())  # rotate and resize axis labels
                                                  # and remove grid lines
```

This plot matrix contains other interesting findings: alcohol has a strong 
negative correlation with density, and a significant positive correlation with 
quality. residual.sugar does not have a significant *linear* correlation with 
quality, suprisingly. The plot for quality and fixed.acidity appears to depict a stronger negative correlation than the Pearson's r of -0.114 given in the plot
matrix. The triangle relationship also occurs for quality and residual.sugar,
and again for quality and citric.acid. In these triangular distributions, the 
quality appears to level off from both directions to a value of 6. 

On further inspection of the matrix, the sulfur dioxide/quality plot is sort of exhibiting the same triangle behavior, until a certain level of sulfur dioxide, 
after which point the quality consistently sits at lower values. 

Based on my entire exploration so far, I want to further explore the 
relationships between quality(and my created binary variables) and 
fixed.acidity; quality(with my created variables) and residual.sugar; and 
quality(with my created variables) and alcohol. It appears that alcohol
has the strongest linear correlation with quality, but I would like to explore
these other relationships involving quality to confirm this. Then I will study 
the supporting variable with the strongest relationship with quality. 

```{r, echo=FALSE, message=FALSE}
# Create scatter plot of alcohol vs. quality, with jitter
ggplot(data = whitewines, aes(x = alcohol, y = quality)) +
    geom_jitter(height = 0.35, width = 0)
```

There does appear to be some correlation between alcohol and quality. I notice 
that the highest quality wines do not have the highest alcohol content. It does
not look like there is much overplotting, but I will add transparency just to
see what happens.

```{r, echo=FALSE, message=FALSE}
# Create scatter plot of alcohol vs. quality, with transparency and jitter
ggplot(data = whitewines, aes(x = alcohol, y = quality)) +
  geom_jitter(alpha=0.4, height = 0.35, width = 0)
```

Adding transparency does reveal more information. While the ranges of existing 
wines at each quality rating are similar, I can see that more wines have higher
alcohol levels as quality increases, at least up until 7. I can see that most 
wines with a rating of 8 have an alcohol level between about 11 and about 13.
Most wines with a rating of 9 have an alcohol rating between 12 and 13. This 
suggests there is a prime range for alcohol levels to be of the best quality. 
This idea is supported by the observation that most wines of lower quality fall
below 11. 

```{r, echo=FALSE, message=FALSE}
# Create box plot of alcohol vs. quality
ggplot(data = whitewines, aes(x = quality, y = alcohol, group = quality)) +
  geom_boxplot()
```

A boxplot sheds more light on the alcohol/quality relationship. From 3 to
9, the mean alcohol level falls until 5, then rises until 9.

```{r, echo=FALSE, message=FALSE}
# Create scatter plot of alcohol vs. rank
ggplot(data = whitewines, aes(x = rank, y = alcohol, group = rank)) +
  geom_boxplot()
```


At the level of rank, the mean alcohol level of bad and average wines are 
similar, but there is more variance for average wines. The mean and bulk of 
alcohol levels of top wines are higher than those of both bad and average wines. 
We again see that the alcohol levels of top wines mostly sit between 11 and 13, 
and the best wines(9) mostly sit between 12 and 13. 

I may look at multivariate relationships involving alcohol and quality, and 
explore what else these top quality wines have in common.

But for now, I will take a closer look at the more peculiar quality
relationships involving fixed.acidity and residual sugar.

```{r, echo=FALSE, message=FALSE}
# Create scatter plot of residual.sugar vs. quality, with jitter
ggplot(data = whitewines, aes(y = quality, x = residual.sugar)) +
    geom_jitter(height = 0.35, width = 0)
```

This distribution of residual.sugar and quality is informative, but there is
still overplotting. Let's apply transparency.

```{r, echo=FALSE, message=FALSE}
# Create scatter plot of residual.sugar vs. quality, with transparency
ggplot(data = whitewines, aes(y = quality, x = residual.sugar)) +
    geom_jitter(alpha=1/10, height = 0.35, width = 0)
```

Transparency helps. What if I apply even more transparency?

```{r, echo=FALSE, message=FALSE}
# Create scatter plot of residual.sugar vs. quality, with more transparency
ggplot(data = whitewines, aes(y = quality, x = residual.sugar)) +
    geom_jitter(alpha=1/18, height = 0.35, width = 0)
```

These scatterplots of the quality/residual.sugar relationship begin to reveal 
more information. There happens to be a somewhat consistent number of wines at
most levels of residual.sugar between 1 and at least 15 for most quality scores.

Let's do a log transform of the x-axis again to see the bimodal distribution of 
residual sugar here.

```{r, echo=FALSE, message=FALSE}
# Create scatter plot of residual.sugar vs. quality, with transparency and a log
# transform of the x-axis
ggplot(data = whitewines, aes(y = quality, x = residual.sugar)) +
  geom_jitter(alpha=1/18, height = 0.35, width = 0) +
  scale_x_log10()
```

Upon applying a log transform to the data, I see that the distributions of 
residual.sugar have the bimodal relationship in common across quality scores: 
plots get denser around 1.5 and again around 10. Therefore, quality does not 
explain the bimodal distribution of residual sugars. 

So why are there two peaks?

Upon research via the internet, I learned about how some white wines are "dry",
and some are "sweet". Dry wines are produced when yeast eats all the sugars and
produces alcohol. Sweet wines are produced when the yeast is stopped before it 
eats all the sugars. Therefore, I expect alcohol and residual sugar to have an
inverse relationship. However, this transformed sugar/quality plot suggests that
quality is not determined by whether a wine is dry or sweet. This suggests 
residual.sugar has a low impact on quality. 

I decided to use my created variables in my study of residual.sugar and quality,
just to see what I find.

```{r, echo=FALSE, message=FALSE}
# Create box plot of residual.sugar vs. rank
ggplot(data = whitewines, aes(x = rank, y = residual.sugar)) +
  geom_boxplot()
```

```{r, echo=FALSE, message=FALSE}
# Output a numerical summary of residual.sugar grouped by rank
by(whitewines$residual.sugar, whitewines$rank, summary)
```

The relationship between rank(groups of wine qualities) and residual.sugar is 
not explained well by a linear model. These boxplots gave me an idea: I wonder 
if a quadratic model would explain the relationship better. 

The boxplots suggest that more sugar does not necessarily mean higher quality, 
which was the same trend for alcohol. 

```{r, echo=FALSE, message=FALSE}
# Create box plot of residual.sugar vs. quality
ggplot(data = whitewines, aes(x = quality, y = residual.sugar, 
                              group = quality)) +
  geom_boxplot()
```

```{r, echo=FALSE, message=FALSE}
# Output a numerical summary of residual.sugar grouped by quality
by(whitewines$residual.sugar, whitewines$quality, summary)

```

Creating the boxplots with quality instead of rank definitely revealed more
information. The trend of how residual sugar level changes as quality increases 
is very non linear. In fact, the highest quality wines (9) have the lowest 
median residual sugar of all the wines! Also, from 7 to 9, residual.sugar 
increases then decreases. This suggests that sugar's impact on quality is low,
which is supported by the very small correlation coefficient of -0.0976. 
So what else is creating the quality boost for these low-sugar wines? I intend 
to look at many multivariate plots of this residual.sugar/quality relationship to 
find out.

Next, I will explore the relationship between fixed.acidity and quality. 

```{r, echo=FALSE, message=FALSE}
# Create a scatter plot of fixed.acidity vs. quality, with transparency
ggplot(data = whitewines, aes(x = fixed.acidity, y = quality)) +
      geom_jitter(alpha=1/10, height = 0.35, width = 0)
```

This scatterplot of quality and fixed.acidity confirms that there is not a 
strong linear correlation between fixed.acidity and quality. The pattern I am
seeing reminds me of residual.sugar/quality: from highest to lowest quality, the 
distribution stretches wider, reaches a peak width, then shrinks again.

For fun, I will look at the relationship between rank and fixed.acidity.

```{r, echo=FALSE, message=FALSE}
# Create a box plot of fixed.acidity vs. rank
ggplot(data = whitewines, aes(x = rank, y = fixed.acidity)) +
  geom_boxplot()
```

```{r, echo=FALSE, message=FALSE}
# Output a numerical summary of fixed.acidity grouped by rank
by(whitewines$fixed.acidity, whitewines$rank, summary)
```

There is not much change in the median or IQR of fixed.acidity as rank 
increases. This boxplot supports the insignificant linear correlation between 
fixed.acidity and quality.

```{r, echo=FALSE, message=FALSE}
# Create a box plot of fixed.acidity vs. quality
ggplot(data = whitewines, aes(x = quality, y = fixed.acidity, group = quality)) +
  geom_boxplot()
```

```{r, echo=FALSE, message=FALSE}
# Output a numerical summary of fixed.acidity grouped by quality
by(whitewines$fixed.acidity, whitewines$quality, summary)
```

Even after breaking the boxplots down further from rank to quality, I see that 
fixed.acidity varies little from poorest to best quality. In addition, the 
poorest and best wines both have higher median fixed.acidity than all the other 
wines! This information supports the idea that fixed.acidity has little impact
on wine quality.


Now that I have explored quality, I have decided that alcohol has the strongest
relationship with quality. Now I will explore how alcohol relates to other 
variables, like residual.sugar(Does it exlain the bimodal distribution? Does
alcohol have a bimodal distribution under a log transform as well?), density, 
and total.sulfur.dioxide.

```{r, echo=FALSE, message=FALSE}
# Create a scatter plot of residual sugar vs. alcohol
ggplot(data = whitewines, aes(x = residual.sugar, y = alcohol)) +
  geom_point()
```

I will create the same plot as above but limit the x axis, so I can study the 
majority of the data more closely.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Create a scatter plot of residual sugar vs. alcohol. Limit the x-axis for 
# closer study
ggplot(data = whitewines, aes(x = residual.sugar, y = alcohol)) +
  geom_point() +
  xlim(0,25)
```

What happens if I do a log transform of the x-axis (residual sugar?)

```{r, echo=FALSE, message=FALSE}
# Create a scatter plot of residual sugar vs. alcohol. Limit the x-axis for 
# closer study, apply transparency and a log transform of the x-axis for 
# closer study
ggplot(data = whitewines, aes(x = residual.sugar, y = alcohol)) +
  geom_point(alpha = .2) +
  xlim(0,25) +
  scale_x_log10()
```

There is somewhat of a pattern here, though it is not obvious. I see that most 
wines fall in what looks like a "sweet" category (higher sugar, less alcohol). 
There appears to be a relatively dense population of wines in the "dry" category
(lower sugar, more alcohol), but it is not clear from this plot. I decided to
increase the transparency for a better look.

```{r, echo=FALSE, message=FALSE}
# Same plot as above with more transparency
ggplot(data = whitewines, aes(x = residual.sugar, y = alcohol)) +
  geom_point(alpha = .06) +
  xlim(0,25) +
  scale_x_log10()
```

From this plot, it is tricky to tell, but I can see the slight negative 
correlation between alcohol and residual sugar; most wines with lower residual
sugar appear to cluster into an alcohol range that is slightly higher than the 
alcohol range for wines with higher residual sugar. 

What if I did a log transform of the y-axis (alcohol)?

```{r, echo=FALSE, message=FALSE}
# Same plot as above with a transform of the y-axis
ggplot(data = whitewines, aes(x = residual.sugar, y = alcohol)) +
  geom_point(alpha = .06) +
  xlim(0,25) +
  scale_x_log10() +
  scale_y_log10()
```

It is just a smidgen easier to see the negative correlation now. I might look at
multivariate plots of this relationship and quality to confirm the finding that
high quality wines can be dry or sweet.

Let's look at alcohol and density.

```{r, echo=FALSE, message=FALSE}
# Create a scatter plot of density vs. alcohol
ggplot(data = whitewines, aes(x = density, y = alcohol)) +
  geom_point()
```

So it is evident from this plot that wines with lower alcohol content are less
dense than higher-alcohol wines. Why is this so? Some brief googling taught me
that alcohol is less dense than water, so this could explain the slight increase
in density as alcohol content decreases; this volume may be substituted with 
more water, making the wine slightly more dense.

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?
My feature of interest, quality, has significant correlation with alcohol.
I found that median alcohol increases as quality increases most of the time, but
the highest quality alcohols tend to stay within a certain range of alcohol 
content. Quality does not appear to have significant linear correlation with 
residual.sugar. Generally, the median residual.sugar increases, peaks, then 
decreases again as quality increases. Fixed.acidity varies little from poorest 
to best quality. In addition, the poorest and best wines both have higher median fixed.acidity than all the other wines! 

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?
I found that there is a negative correlation between alcohol and residual.sugar,
though it is not obvious until you apply transparency and log transformations to
the data. I also saw a very strong negative correlation between alcohol and 
density. 

### What was the strongest relationship you found?
The strongest relationship I found was between residual.sugar and density. The
plot matrices I created show that the correlation coefficient of residual.sugar
and density is 0.839.

# Multivariate Plots Section

For my multivariate plots, I will explore alcohol and quality with a third 
variable, like residual.sugar, to understand what else top quality wines have in
common.

To assist with these plots, I installed the colorRamps package and created a 
palette(though I did not make much use of them).

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# load colorRamps package
library(colorRamps)
```


```{r, echo=FALSE, message=FALSE, warning=FALSE}
# assign hex colors to a vector, cbPalette 
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
```

```{r echo=FALSE, Multivariate_Plots}
# create a scatter plot of residual.sugar vs. alcohol, faceted by quality
ggplot(data = whitewines, aes(x = residual.sugar, y = alcohol)) +
    geom_point(alpha = .2) +
  scale_x_log10() +
  facet_wrap(~quality)
```

This faceted scatterplot enables me to directly visualize what I had discovered 
from my bivariate exploration. When I facet by quality, I find that wines of all 
quality scores have have a residual.sugar level that varies from 1 to at least 
about 10. However, you can see the cloud of points rise upward to higher alcohol 
levels as quality increases, until you reach the highest quality wines, most of
which land in the 12-14 range. However, the highest quality wines (8,9) still
have some variety in their range of possible alcohol levels. Is there another
variable that distinguishes the top wines of lower alcohol content from those of
the more common alcohol range? Let's explore.

I will do so by experimenting with different second variables in the x-axis and 
keeping quality as the third variable indicated by color in multiple 
individual scatterplots.

First, I will look at pH.

First I turned quality into a factor.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Convert quality from int to factor
whitewines$quality = factor(whitewines$quality)
```

Then I replaced residual sugar with pH in the scatterplot:

```{r, echo=FALSE, message=FALSE}
# create a scatter plot of pH vs. alcohol, with transparency, faceted by quality
ggplot(data = whitewines, aes(x = pH, y = alcohol)) +
    geom_point(alpha = .2) +
  facet_wrap(~quality)
```

pH does very, very slightly appear to differentiate top wines by alcohol 
content. I can see that many high-alcohol wines sit in the same pH range 
as the lower-alcohol wines, but the lowest-alcohol wines tend to stay at lower
pH ranges. 

Therefore, acidity appears to have some involvement in what makes lower-alcohol
wines top quality.

what about citric.acid?

```{r, echo=FALSE, message=FALSE}
# create a scatter plot of citric.acid vs. alcohol, with transparency, faceted 
# by quality
ggplot(data = whitewines, aes(x = citric.acid, y = alcohol)) +
    geom_point(alpha = .2) +
  facet_wrap(~quality)
```

Citric acid does not appear to tell us much. The ranges of citric acid levels 
for top quality wines do not vary much as alcohol level increases.

What about fixed.acidity(most acids in wine)?

```{r, echo=FALSE, message=FALSE}
# create a scatter plot of fixed.acidity vs. alcohol, with transparency, faceted 
# by quality
ggplot(data = whitewines, aes(x = fixed.acidity, y = alcohol)) +
    geom_point(alpha = .2) +
  facet_wrap(~quality)
```

As with pH, I see a very slight trend in fixed.acidity's relationship with 
alcohol. For the top quality wines (8,9), The lowest-alcohol wines (below 10) 
tend to have higher fixed acidity levels than most higher-alcohol wines. The 
trend is more obvious for the plot of wines with a quality score of 9, but it is visible with 8's as well, thanks to transparency.

What about volatile.acidity(acetic acid)?

```{r, echo=FALSE, message=FALSE}
# create a scatter plot of volatile.acidity vs. alcohol, with transparency, 
# faceted by quality
ggplot(data = whitewines, aes(x = volatile.acidity, y = alcohol)) +
    geom_point(alpha = .2) +
  facet_wrap(~quality)
```

There seems to be more complex relationship between the volatile acidity and 
alcohol content of top wines. It is reminiscent of the relationship between 
alcohol and quality for all wines. The volatile acidity of lower-alcohol wines
tends to be lower, but the lowest-alcohol wines sit in a particular range of 
volatile acidity that is higher than some. This trend is again more obvious for
wines with a score of 9.

Let's look at fixed.acidity and volatile.acidity together in the faceted 
scatterplot. I will use the same plot as above, but replace alcohol with 
fixed.acidity and use alcohol as color.

```{r, echo=FALSE, message=FALSE}
# Create a scatter plot of volatile.acidity vs. fixed.acidity, 
# colored by alcohol, with transparency and reduced point size, faceted by 
# quality
ggplot(data = whitewines, aes(x = volatile.acidity, color = alcohol, y=fixed.acidity)) +
    geom_point(alpha = .7, size =.98) +
  facet_wrap(~quality) +
  # add purple-orange color theme for alcohol, with a legend as a color guide
  scale_color_gradient(guide="legend", low = "purple", high = "orange",
                       aesthetics = "color") 
```

This plot helps me see that there is a pattern involving acidity and 
alcohol content for top quality wines. I will subset the dataset to just top quality wines to help me zoom in on this relationship.

```{r, echo=FALSE, message=FALSE}
# Similar plot as above, but only plot wines with a quality of 8 or 9. No size
# or transparency arguments this time
ggplot(data = subset(whitewines, quality == 8 | quality == 9), aes(x = volatile.acidity, color = alcohol, y=fixed.acidity)) +
    geom_point() +
  facet_wrap(~quality) +
  scale_color_gradient(guide="legend", low = "purple", high = "orange",
                       aesthetics = "color")
```

Eureka. I can see that among top wines, the lower-alcohol wines tend to have 
hgiher fixed acidity and lower volatile acidity. This is evident by how the more
purple points, which represent lower alcohol content, tend to reside in the 
upper left quadrant of the points. 

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?
The alcohol-residual.sugar-quality relationship allowed me to notice that though 
the alcohol-quality relationship is significantly linear, there are some wines
with very low alcohol content, yet high quality. This led me to explore what 
else made these low-alcohol high quality wines stand out. 

### Were there any interesting or surprising interactions between features?
I found that top quality wines with lower alcohol levels tend to have both 
higher fixed acidity levels and lower volatile acidity levels to an extent. 

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.
I did not create a model in this exploration.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, message=FALSE, Plot_One}
# Plot histogram of alcohol, with color changes, main title, and x-axis label
ggplot(whitewines, aes(x=alcohol)) +
  geom_histogram(color = "blue", fill="orange") +
  ggtitle("Distribution of Alcohol Content") +
  xlab("percent alcohol")

```

### Description One
In my univariate exploration, I looked at the distribution of alcohol content 
for all wines in the dataset. Most wines sit between 9 and 12% alcohol, with the
most common alcohol content being about 9.2%.

### Plot Two
```{r echo=FALSE, Plot_Two}
# Create box plot of alcohol vs. quality, with color changes, main title, and x-
# and y-axis labels
ggplot(data = whitewines, aes(x = quality, y = alcohol, group = quality)) +
  geom_boxplot(color="purple", fill="yellow") +
  ggtitle("Alcohol Content vs. Quality") +
  ylab("percent alcohol") +
  xlab("quality score")
```

### Description Two
In my bivariate exploration, I found that alcohol has the strongest linear 
correlation with quality. The above plot allowed me to visualize the 
relationship and find out that while the median alcohol content increases as you
approach higher levels of quality, most of the highest quality wines have an alcohol content between 11 and 13%, which is actually lower than the alcohol content of some other wines. 

### Plot Three
```{r echo=FALSE, Plot_Three}
# Create a scatter plot of volatile.acidity vs. fixed.acidity, colored by 
# alcohol, faceted by quality, with purple-orange color theme for alcohol, 
# legend for color theme, main title, x- and y-axis labels, and legend title
ggplot(data = subset(whitewines, quality == 8 | quality == 9), aes(x = volatile.acidity, color = alcohol, y=fixed.acidity)) +
    geom_point() +
  facet_wrap(~quality) +
  scale_color_gradient(guide="legend", low = "purple", high = "orange",
                       aesthetics = "color") +
  ggtitle("Acidity, Alcohol for Top Quality Wines (Score of 8 or 9 out of 10)")+
  xlab("volatile acidity (g/dm^3)") +
  ylab("fixed acidity (g/dm^3)") +
  guides(color=guide_legend(title = "alcohol (%)"))

```

### Description Three
In my multivariate exploration, I studied to find out why some wines have 
alcohol content that is less than 11% but are still top quality. The above plot
gives some insight: low-alcohol top-quality wines tend to have lower volatile acidity and higher fixed acidity than other top quality wines with typical
alcohol content.

------

# Reflection

This exploration was ultimately a successful one because of the insights I was 
able to gain in each phase, from univariate to multivariate. The data is limited
in that I had less data for wines of top quality (scores of 8-9) or poor quality (scores of 3-4), so my ability to look at differences within those ranks was
limited. The dataset, though it has sufficient features to enable a meaningful 
exploration of differences within ranks of wine, might be less detailed when it comes to features like fixed acidity; if there are multiple acids 
involved in this feature, it might be useful to break this variable down for a
deeper exploration.
What went well were my strategies for narrowing down my features of interest. My method of picking features from my first plot matrix to use in my second plot matrix led to me finding some interesting relationships that increased my understanding of wine quality. My choices of plots and aesthetics for studying relationships also went well most of the time. For example, selecting both scatterplots and boxplots to explore the relationship between alcohol and 
quality was a good decision, as each plot revealed unique and useful 
information.
I ran into a few road blocks as I picked aesthetics for more complicated plots.
In my multivariate plots, I tried to color by quality at first, but this made it harder to visualize other features. I eventually decided to facet by quality, 
which made visualizing other features easier. My color scheme choices for my multivariate plots were also a roadblock; the default single hue gradient made 
it very difficult to visualize differences in color. It took a few attempts to
settle on the orange/purple theme that works well.
There is quite a bit of work I could do with this dataset in the future. One
idea is to explore the more complicated relationships between quality and other
features like citric.acid and residual.sugar: Why does quality level off at a
value of 6 as these variables increase? I could attempt to explain these relationships with a quadratic model line instead of a linear model line.
More ideas would be to try to understand the unusual gap in the scatterplot of total.sulfur.dioxide and quality, or the significant correlation between density and quality. Lastly, I could have studied other supporting variables besides acidity to determine what else makes low-alcohol top-quality wines different
than top wines with typical alcohol content for that score.